With this thesis, I presented Rust-Stainless, a formal verification tool for Rust
based on the Stainless verifier framework. Building on a solid foundation of
existing software and infrastructure, I added many features to the tool and thus
extended its usability and expressiveness significantly. Along with smaller
language features the most important additions are immutable references and
boxes, the trait to  type class translation and, first and foremost, the
mutability translation. Further, this project includes the theoretical reasoning
behind the correctness of the mutability translation as well as numerous
bugfixes and additions to both Rust-Stainless and Scala Stainless.

The development went hand-in-hand with testing and evaluating the tool on more
and more complex Rust examples. As an intern at Informal Systems, I had access
to experienced Rust developers who advised me on the choice of features to
support. Chapter \ref{evaluation} shows how the tool was evaluated against some
code of Informal Systems and it also demonstrates that the translations
performed by our tool are very fast, in comparison to the time it takes to
verify the code.

By combining Rust's borrow checking with Stainless's verification,
Rust-Stainless gains an unimagined expressiveness. Thanks to Rust features like
mutability and \lstinline!mem::replace! along with Stainless's new
\lstinline!freshCopy! primitive, the tool can now process code that would fail
Stainless's aliasing restrictions in Scala but is proven memory safe by the Rust
compiler. The support for in-place updates with mutability unlocks the
possibility of verifying actual, efficient Rust code with Stainless, like a
search tree -- without using duplicate functional ghost structures for
verification.


\paragraph{Extraction}

There are still some indispensable language features that Rust-Stainless needs
to support before it can extract idiomatic Rust in all its facets. The primary
ones are loops, arrays, and closures. However, the biggest limitation of the
tool  is the one-crate-limitation inherited from the compile model of Rust.
Solving it would solve many related problems. Instead of supporting vectors, one
could add sensible contracts for vectors and iterators. More generally, with the
ability to attach contracts to crate-external items, Rust-Stainless could
finally  use standard library items like traits for equality, ordering and
hashing. Additionally, we could provide real implementations in the Stainless
crate instead of just providing bindings and replacing them with generated
implementations. This would also enable the use of verified data structures like
the list-map.

\paragraph{Mutability}

The mutability translation presented in this thesis can theoretically translate
all Rust mutability to Scala, but the current implementation is limited by the
imperative phase of Stainless. With the current backend, our tool has probably
maximised the features that it can support. Even if the imperative phase was
bug-free, there are still the two drastic limitations of recursive functions
returning mutable references and mutable references in mutable variables.

The promising solution to that problem would be to use the new fully imperative
phase that is currently developed for Stainless. That phase does not have
aliasing restrictions but rather its own notion of heap references. Hence, one
could remove some of the Stainless specific changes to the translation, thus
making it simpler.  Additionally, I implemented the mutable cell synthesis in
such a way that it would be easy to target the new heap reference cells instead.


\paragraph{Counter-Examples}

Even if Rust-Stainless seems to be the only formal verification project for Rust
that translates from the THIR, instead of the lower-level MIR or LLVM IR, this
thesis project showed the decision to be productive. Most Rust constructs can be
translated in a one-to-one fashion to Stainless AST. Even more so,
Rust-Stainless could turn  that decision into a virtue by becoming the first
Rust verification tool that is counter-example complete! As Stainless is
counter-example complete, it suffices to translate the Scala counter-examples
back to Rust.



\paragraph{New Applications}

The ability of verifying high-level properties on Rust code sparks numerous
ideas for new applications of Rust-Stainless. For example, if software needs to
be verified and very efficient, one can write it in Scala, verify it with
Stainless and then use Stainless's C-code-generation feature. The generated code
is memory safe but of course not as optimised as hand-written C. A much simpler
approach would be to use Rust-Stainless  to directly verify the efficient Rust
code, given that it supports all the used language features.

Another idea is to use Rust-Stainless to verify implementations that are
compiled to \emph{Web Assembly (WASM)}. For example, the IBC relayer of Informal
Systems reads policies that govern its functioning from WASM. Rust-Stainless
could  be used to guarantee certain properties on these policies, e.g. that no
rules are conflicting.


\hfill \break \noindent  In summary, with this thesis and the implementation of
our tool, Rust-Stainless, I showed that the combination of Rust and Stainless is
very powerful. While the tool has its limitations, it also offers many promising
paths for future work and  it can already be used to verify real-world Rust
code, making systems more correct and safer.
