This chapter presents how Rust-Stainless is tested and gives some examples of
code that the tool can verify. Furthermore, it will decompose the running time
of a typical execution into its subparts and finish by presenting a code example
related to distributed systems implemented by Informal Systems and verified with
Rust-Stainless.

\section{Benchmarks}

The Rust-Stainless repository contains a test suite of 67 passing (positive) and
12 failing (negative) test code examples of in total around 3000
lines.\footnote{Found under \texttt{stainless\_frontend/tests/[pass|fail]} on
the \texttt{mutable-cells} branch.} The suite is run for every pull request and
every commit to the \texttt{master} branch.

To quantitatively evaluate Rust-Stainless, I collected some statistics of all
the positive test examples, displayed in \autoref{measurements}. Only the
positive examples were used for the measurements because the negative examples
may fail at different stages of the extraction or verification pipeline. Thus,
the time measurement is only representative for the positive examples which
complete the verification pipeline. The used benchmarks amount to 2214 lines of
Rust code (LoC), Stainless generates in total 475 verification conditions (VCs)
and the accumulated running time is 5.1 minutes (Total). Note that the test
suite normally (when run with \texttt{cargo test}) runs multiple tests in
parallel and thus completes in less than 2 minutes on the same
machine.\footnote{All the tests were run on a MacBook Pro with a 2.8~GHz
Quad-Core Intel Core i7 and 16~GB of RAM.}

It is instructive to decompose the running time into its different parts. As
described in \autoref{sec:extraction-overview}, the frontend first needs to
detect some standard library items before it can translate the user crate. On
average, detection takes $0.17 \pm 0.01$~s or equivalently 4~\% of the running
time. The actual translation is very fast, especially for the rather short
examples that account for the majority of the test suite. It takes only 0.4~ms
or less than 0.01~\% of the running time on average. By far the largest part of
the running time is not spent in the frontend itself but rather in the JVM
backend. Verification time accounts for on average 95~\% of the total time. The
rest is made up of starting, serialising and reporting. This shows that the
translation time is inferior to the verification time by multiple orders of
magnitude which is not surprising, given that verification is much more complex
than translation.

\input{measurements/measurements.tex}


\section{Code Examples}

After having quantitatively measured our tool, it is time to qualitatively
evaluate what it is capable of and to that end present some code examples. The
first three examples stem from the test suite of Rust-Stainless and are included
in the measurements from above. The fourth example is the running example from
\autoref{background} and the last one is an implementation by Informal Systems
used in the \texttt{tendermint-rs} repository.

\subsection{Test Suite Examples}

\paragraph{Insertion Sort}

The first example is \autoref{lst:insertion-sort}, an implementation of
\emph{insertion sort} on the already mentioned functional linked-list. The
recursive data type is made possible by the support for boxes. This example was
translated from an equivalent Scala example in Stainless's test suite, therefore
it shows a functional way of writing Rust; there is no mutation in the entire
example. In idiomatic Rust, one would probably use a vector instead of a
linked-list. The example further showcases the use of implementation blocks and
specs on methods. For proving termination of the recursive implementation it is
crucial to add the new \lstinline!measure! attribute. Also note that the
\lstinline!Option! type is the one from the standard library.

Because the example is completely functional, it serves perfectly to evaluate
whether the  mutability translation preserves functional code -- despite the
mutable cell encoding. To test that, consider the \lstinline!sorted_insert!
method, displayed in \autoref{lst:sort-ins-rs}, and its translation as it is
submitted to Stainless, shown in \autoref{lst:sort-ins-1}. The translation is
almost identical. The only changes are the mutable cell patterns in the match,
the \lstinline!freshCopy! around the return values, and the erasure of the
boxes. This is thanks to the optimisations of \autoref{optimisations}. For
example, the function arguments are not wrapped in mutable cells because they're
immutable.

\noindent\begin{minipage}[t]{.43\textwidth}
\begin{lstlisting}[
  language=Rust,
  basicstyle=\footnotesize\ttfamily,
  label={lst:sort-ins-rs},
  caption={Completely functional method of \autoref{lst:insertion-sort}.}
]
#[pre(self.is_sorted())]
#[measure(self)]
#[post(
  ret.size() == self.size() + 1 &&
  ret.is_sorted() &&
  ret.contents().is_subset(
    &self.contents().insert(e)
  ) &&
  self.contents().insert(e).is_subset(
    &ret.contents()
  )
)]
pub fn sorted_insert(
  self,
  e: i32
)-> List<i32> {
  match self {
    List::Cons(head, tail)
      if head <= e => List::Cons(
        head,
        Box::new(tail.sorted_insert(e))
      ),
    _ => List::Cons(e, Box::new(self)),
  }
}
\end{lstlisting}
\end{minipage}\hfill
\begin{minipage}[t]{.55\textwidth}
\lstinputlisting[
  language=Scala,
  basicstyle=\footnotesize\ttfamily,
  label=lst:sort-ins-1,
  caption={Translation of \autoref{lst:sort-ins-rs}.}
]{code/sorted_insert_1.scala}
\end{minipage}

To go a step deeper, consider the final encoding of the function, displayed in
\autoref{apx:insertion-encoding}. Stainless submits this encoding to its solver,
Inox, after having executed all its transformations. In particular, after the
imperative phase has transformed all mutable fields like the one of the mutable
cells to functional code. The encoding contains refinement assertions and
multiple casts to use the fields of the ADT directly, e.g.~\lstinline!._1!,
instead of using the matched bindings. Otherwise, it does not introduce
unnecessary complexity to deal with mutable cells. Thus, it can be concluded
that the mutability translation preserves the verifiability of functional code
-- at least as long, as no type classes are involved that may cause trouble with
the refinement lifting.

\paragraph{Type Class}

The next example in \autoref{lst:type-class} is also completely immutable and
demonstrates the use of a trait for equality (\lstinline!Equals!) with two
implementations. Note that one would usually use the traits from the standard
library (\lstinline!Eq! and \lstinline!PartialEq!) and let the macro derive
implementations for them. Here however, equality is implemented in the program
because it serves as a good example for contracts on traits. The trait contains
three laws corresponding to the three properties of the equivalence relation.

The implementation for \lstinline!i32! is trivial except that it needs to
dereference the two operands. This is to force the compiler of using the
primitive comparison operator instead of  \lstinline!PartialEq::eq!. The
linked-list implementation (\autoref{lst:equals-list}) is a good example of a
trait bound (\lstinline!T: Equals!) and two type class method calls that will be
resolved in the translation.

\begin{lstlisting}[
  float=htb,
  language=Rust,
  caption={Equality implemented for the linked-list.},
  label={lst:equals-list}
]
impl<T: Equals> Equals for List<T> {
  fn equals(&self, other: &List<T>) -> bool {
    match (self, other) {
      (List::Nil, List::Nil) => true,
      (List::Cons(x, xs), List::Cons(y, ys)) => x.equals(y) && xs.equals(ys),
      _ => false,
    }
  }
  ...
}
\end{lstlisting}

Lastly, note that the three laws need to be reimplemented in the list
implementation. This is due to Stainless not being able to recursively prove the
properties without some guidance by the programmer. However, the same is
necessary in an equivalent Scala example. Thus, this is not a limitation of the
Rust frontend but of Stainless itself. On the other hand, the law
implementations show another construct that the frontend is able to translate:
static method calls like \lstinline!Self::law_transitive(xs, ys, zs)! are
translated to calls on the correct type class instance.

\paragraph{Local Mutability}

Turning to mutability, the benchmark in \autoref{lst:local-clone} is available
on the \texttt{mutable-cells} branch. It features a struct for which the macro
derives a \lstinline!Clone! instance. In the function, the struct is created,
cloned and mutated. The assertions ensure that the cloned instance is not
changed as well by the mutation. The mutability translation  will replace the
\lstinline!clone! call with \lstinline!freshCopy!.

\begin{lstlisting}[
  float=bht,
  language=Rust,
  caption={Local mutation and cloning of a simple struct.},
  label=lst:local-clone
]
#[derive(Clone)]
pub struct S(i32);

pub fn main() {
  let mut a = S(1);
  let b = a.clone();
  a.0 = 10;
  assert!(a.0 != b.0)
}
\end{lstlisting}


\paragraph{Mutable References}

The running example (\autoref{running-example}) from \autoref{background} is
equivalent to the most complicated mutability benchmark on the
\texttt{mutable-cells}
branch\footnote{\url{https://github.com/epfl-lara/rust-stainless/blob/mutable-cells/stainless_frontend/tests/pass/mut_ref_borrow_12.rs}}
with the addition of a trait. In particular, it shows how a mutable reference is
matched upon and then a sub-reference is returned. The sub-reference is later
used to alter the container struct. In summary, the example requires support for
mutable borrows, pattern matching on mutable references, passing mutable
references as values, and struct mutation via references.

In the translation of \lstinline!get_mut_by_id!
(\autoref{lst:mut-ref-translated}), one can observe how the pattern match binds
the mutable cell object \lstinline!v! inside the tuple ADT instead of its value.
Furthermore, the evidence argument responsible for the trait method,
\lstinline!ev0.id! on line 8, is added as parameter.

\begin{lstlisting}[
  float=htb,
  language=Scala,
  caption={Translation of the \lstinline!get_mut_by_id! method from \autoref{running-example}.},
  label={lst:mut-ref-translated}
]
def get_mut_by_id[K @mutable, V @mutable](
  self: MutCell[Container[K, V]],
  id: Long,
  ev0: Id[K] @evidence
): Option[V] = {
  self.value.pair match {
    case MutCell(Some(MutCell(Tuple2(k, v))))
      if ev0.id(k.value) == id => Some[V](v)
    case _ => None[V]()
  }
} ensuring {
  (ret: Option[V]) => is_empty[K, V](self.value) ==> ret match {
    case None() => true
    case _ => false
  }
}
\end{lstlisting}

\subsection{Peer-List Implementation}
\label{peerlist}

A principal motivation of this thesis project was to verify real-world code
examples of Informal Systems. The \lstinline!PeerList! data structure seemed
like the perfect benchmark because it already had runtime-checked invariants
defined by Romain Rüetschi. Thus, a primary driver of this thesis was to verify
a version of the data structure that was as close to the original code as
possible.

Thanks to the mutability translation and especially in-place updates, I achieved
that goal adequately, found a problem in the original specification and was able
to strengthen the postconditions of some methods. The verified version of the
code is in \autoref{apx:peerlist} and by comparing it with the original code,
one can see that there are indeed no large changes.\footnote{The original
version can be found here:
\url{https://github.com/informalsystems/tendermint-rs/blob/d8e18c647cd8695d16610c4292b15ec6d1b45fbc/light-client/src/peer_list.rs}.}

The peer-list is a data structure that keeps track of different nodes in a
distributed system context. Each node has a current value, recorded in the
\lstinline!values! map, and each node is either the primary node of the system,
a trusted witness, a simple but working full-node, or considered as a faulty
node. These states are recorded with sets in the struct and the
\lstinline!invariant! method returns true if the sets are in a valid
configuration. For example, they all  need to be disjoint as each node must be
in exactly one state. The two methods of interest are
\lstinline!replace_faulty_witness! and \lstinline!replace_faulty_primary! that
both change the state of some of the nodes.

The only notable but crucial change from the original code is that instead of
the standard \lstinline!HashMap! and \lstinline!HashSet!, the verified code uses
two custom implementations, \lstinline!ListMap! and \lstinline!ListSet!. I
implemented these two specially for the benchmark. They are in a separate module
in the verified crate (\autoref{apx:listmap}). The need for list-backed set and
map implementations arose because some methods get an element from the
collections with \lstinline!.iter().next()! in the original code. Neither does
Rust-Stainless support iterators, nor do the map and set from Stainless support
element retrieval. Therefore, I created the list-backed collections and added a
\lstinline!first! method. Ideally, the two list-collections should be in the
Stainless crate and be part of the library. If that was the case, the peer-list
example would only have to import the Stainless crate and change  the collection
types, to be verifiable. Unfortunately, the one-crate-limitation (cf.
\autoref{impl-limitations}) currently forces us to have the collections in the
user crate.

Nonetheless, verifying the peer-list implementation with
\passthrough{\lstinline!cargo stainless!} allowed me to improve its
specifications. First, I found a problem: the precondition that the invariant
must hold before the two methods was not set in the code. Furthermore, I could
strengthen the postconditions of both methods to also prove that the methods
perform the correct changes. For example, \autoref{lst:postcondition} shows the
improved postcondition of \lstinline!replace_faulty_primary!. If the method
succeeds and returns an \lstinline!Ok! result, the \lstinline!self! not only has
to satisfy the invariant but the faulty primary given as argument also needs to
be in the faulty nodes. This prevents the method from doing nothing which would
also uphold the invariant. Note also the use of the \lstinline!old(&self)!
helper to use the value of \lstinline!self! before the function.

\begin{lstlisting}[
  language=Rust,
  caption={Postcondition of the \lstinline!replace_faulty_primary! method.},
  label={lst:postcondition}
]
#[post((matches!(ret, Ok(_))).implies(
  Self::invariant(&self)
    && old(&self).primary != self.primary
    && self.faulty_nodes.contains(&old(&self).primary)
    && old(&self).witnesses.contains(&self.primary)
))]
\end{lstlisting}

The tool is also capable of verifying another example from Informal Systems. The
code stems from an \emph{inter-blockchain communication protocol (IBC)} handler
implementation and was rewritten to be immutable and self-contained because it
was used on an earlier version of the tool. The current implementation still
verifies the example.\footnote{The PR with the example can be found here:
\url{https://github.com/informalsystems/ibc-rs/pull/759}.}

\hfill \break \noindent With five different examples showcasing each their set
of language features, this chapter showed what Rust-Stainless can currently
process. It became clear that no matter how complicated the translations in the
frontend are, the running time is dominated by the verification time of the
backend. For an overview of the tool's limitations, see
\autoref{impl-limitations}. The next chapter summarises research  work in
related areas like ownership systems and formal verification.
